---
title: "INLA" # title of the notebook
author: "Virginie Sonnet \nvirginie_sonnet@uri.edu \nhttps://github.com/VirginieSonnet" # name, email and github with links
date: "`r format(Sys.Date(), '%B %dth, %Y')`" # date of today (change th if today is 1st, 2nd or 3rd)
output:
  # specific format from the R package "rmdformats"
  rmdformats::readthedown:
    lightbox: TRUE # allows to click on an image to zoom in the HTML document
    highlight: pygments # color code for code (default is also a good option)
    toc_depth: 5 # level of headers to include in the table of contents 
    # code_folding: "hide" # do not show the code (can access it with a button)
    # use_bookdown: TRUE # automatic section numbering 
    # toc_float: TRUE # for regular markdowns, allow the toc to not be on top but on the side
---

<style>
body {
font-family: "LM Roman"}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

```{r}
rm(list=ls())
gc()
```

```{r}
# data manipulation
library(tidyverse) # data manipulation
library(lubridate) # dates
library(imputeTS) # missing values imputation 
library(TSA) # harmonic components 

# fitting
library(INLA) # install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
# inla.update(testing=T)

# parallelization 
library(parallel)
detectCores()

# plotting
source("INLA_brinla_updated.R")
library(extrafont)
loadfonts()
```

## Introduction

This code shows how to run an INLA model on morphological variables, using environmental variables as predictors. 



## Data 

We first need to load the morphological coordinates averaged per hour and the environmental data. We only keep temperature, light and salinity as regressors since they have the hourly resolution. 

```{r}
# morpho
pca <- read_csv("../data/pca_coord_summary.csv") %>% 
  select(date, ends_with("mean") & !starts_with("conc")) %>% 
  rename_at(vars(ends_with("_mean")), ~str_remove(.,"_mean"))

# environment
envt <- read_csv("../data/NBay_hourly.csv", guess_max=3000) %>% 
  select(date, temp, sal, light)

# join
data <- left_join(pca, envt) %>% 
  # convert to EST time zone
  mutate(date = with_tz(date, tzone = "EST")) 
```

Now that we have our dataset, we want to scale our regressors so that their coefficients are directly comparable.  

```{r}
data <- data %>% 
  mutate_at(vars(-all_of(c("date","PC1","PC2","PC3","PC4"))),  ~(scale(.) %>% as.vector)) 
```

## INLA 

We first set-up our different vectors. 

```{r}
# Vectors of values 
# ---
temp = as_vector(data$temp) # temperature
sal = as_vector(data$sal) # salinity
light = as_vector(data$light) # light


# Corresponding length vectors 
# ---
N = nrow(data)
t = seq(1, N) # temperature
s = seq(1, N) # salinity
l = seq(1, N) # light

j = seq(1,N) # for seasonal component
ind1 = seq(1,N) # for dynamic intercept
```

Now we need to construct our regression formula with: 

* dynamic intercept 
* seasonal component
* temperature
* salinity 
* light 

```{r}
# Formula 
# ---
formula = y ~ f(ind1, model = "rw1", constr=T) +
  f(j, model="seasonal", season.length=24) +
  f(t,temp,model="rw1",constr=F) +
  f(s,sal,model="rw1",constr=F) + 
  f(l,light,model="rw1", constr=F) - 1
```

Calling INLA can be done by hand or you can loop or use a function over your 4 different variables. Here we use a function, `morpho_inla` that extracts the morphological variable and run the INLA model on it using the formula we created before. We store all of the main results in an R object (we used to store the whole INLA model but files often get corrupted). 

If you have several cores that are free on your computer, you can run the model in parallel. However, note that `mc.cores > 1` is only supported on Mac, not on Windows.  

```{r}
# morphological variables 
var <- c('PC1', "PC2" ,"PC3","PC4")


# function to generate and save the results 
morpho_inla <- function(var){
  y <- as_vector(data[[var]])
  res = inla(formula=formula,data=data.frame(y,ind1,l,t,s,j),family="gaussian",
             control.predictor=list(compute=TRUE), 
             control.compute = list(dic = TRUE, waic = TRUE,mlik=TRUE,cpo=TRUE),
             control.inla = list(strategy = "gaussian",int.strategy = "eb"), 
             num.threads = 1,verbose=TRUE)
  r <- list(names.fixed = res$names.fixed,
            summary.fitted.values = res$summary.fitted.values,
            marginals.fitted.values = res$marginals.fitted.values,
            marginals.fixed = res$marginals.fixed,
            summary.fixed = res$summary.fixed,
            summary.random = res$summary.random,
            misc = res$misc,
            waic = res$waic$waic,
            dic = res$dic$dic)
  rm(res)
  saveRDS(r, file=paste("../data/inla_",var,".rds",sep=""))
} 


# parallelization 
results <- mclapply(var, morpho_inla, mc.cores = 4) # leave 2 cores for other processes
```

## Graphs 

Now we can plot the random effects (temperature, light, salinity and the dynamic intercept). This script uses a modified version of the functions of the package **brinla** (https://github.com/julianfaraway/brinla). 

```{r}
labels <- c("ind1"="intercept","j"="seasonal","s"="salinity","t"="temperature","l"="light")

# load the data 
pc1.mod <- readRDS("../data/inla_pc1.rds")
pc2.mod <- readRDS("../data/inla_pc2.rds")
pc3.mod <- readRDS("../data/inla_pc3.rds")
pc4.mod <- readRDS("../data/inla_pc4.rds")
``` 

```{r}
plot_random_effects(
  pc1.mod, date=data$date, date_breaks="4 months", label=labels, 
                           title="PC1 - Size",legend=FALSE) + theme(axis.line.x=element_line(linetype="dotted",color="grey90",size=0.75),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_line(color=" grey98"))

ggsave(filename="F9_INLA1.tiff", path = "../figures/", device='tiff', dpi=300)
```

```{r}
plot_random_effects(
  pc2.mod, date=data$date, date_breaks="4 months", label=labels, 
                           title="PC2 - Width",legend=FALSE) + theme(axis.line.x=element_line(linetype="dotted",color="grey90",size=0.75),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_line(color=" grey98"))
ggsave(filename="F9_INLA2.tiff", path = "../figures/", device='tiff', dpi=300, 
       width=4.41, height=5.69, units="in")
```

```{r}
plot_random_effects(
  pc3.mod, date=data$date, date_breaks="4 months", label=labels, 
                           title="PC3 - Regularity",legend=FALSE) + theme(axis.line.x=element_line(linetype="dotted",color="grey90",size=0.75),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_line(color=" grey98"))
ggsave(filename="F9_INLA3.tiff", path = "../figures/", device='tiff', dpi=300,
       width=4.41, height=5.69, units="in")
```

```{r}
plot_random_effects(
  pc4.mod, date=data$date, date_breaks="4 months", label=labels, 
                           title="PC4 - Chains",legend=FALSE) + theme(axis.line.x=element_line(linetype="dotted",color="grey90",size=0.75),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.x = element_line(color=" grey98"))
ggsave(filename="F9_INLA4.tiff", path = "../figures/", device='tiff', dpi=300,
       width=4.41, height=5.69, units="in")
```





