---
title: "PCA" # title of the notebook
author: "Virginie Sonnet \nvirginie_sonnet@uri.edu \nhttps://github.com/VirginieSonnet" # name, email and github with links
date: "`r format(Sys.Date(), '%B %dth, %Y')`" # date of today (change th if today is 1st, 2nd or 3rd)
output:
  # specific format from the R package "rmdformats"
  rmdformats::readthedown:
    lightbox: TRUE # allows to click on an image to zoom in the HTML document
    highlight: pygments # color code for code (default is also a good option)
    toc_depth: 5 # level of headers to include in the table of contents 
    # code_folding: "hide" # do not show the code (can access it with a button)
    # use_bookdown: TRUE # automatic section numbering 
    # toc_float: TRUE # for regular markdowns, allow the toc to not be on top but on the side
---

<style>
body {
font-family: "LM Roman"}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

```{r}
library(tidyverse)
library(lubridate)
library(data.table)
library(factoextra)
library(grid)
library(extrafont)
library(wesanderson)
library(svglite)
loadfonts()
library(hrbrthemes)
```

## Introduction

This code looks at the dominant phytoplankton groups each week. 

## Phytoplankton data 

A general phytoplankton classification is stored in the Mouw lab database (classifier 7). A weekly sum of the counts for each species is stored in the file *gsoIFCB_date+classnames+counts_weekly.csv*.

We remove all non-identieid or non-living classes. 

```{r}
data_week <- read_csv("../data/gsoIFCB_date+classnames+counts_weekly.csv")

# remove non-identified or non-living classes 
data_week_filter <- data_week %>% 
    filter(!(class %in% c("unclassified","bad","bead","bubble")))
```

We calculate the percentage that each class represents during each week and which one is the class with the highest percentage. 

```{r}
# calculate the total concentration per week
counts_week_filter <- data_week_filter %>% 
    group_by(date) %>% 
    summarize(tot_counts = sum(counts))

# add to the table and calculate the percentage for each group
data_week_filter <- data_week_filter %>% 
    left_join(counts_week_filter) %>% 
    mutate(perc = counts*100/tot_counts)

# main classes in percentage
max_week_filter_perc <- data_week_filter %>% 
    group_by(date) %>% 
    filter(perc == max(perc)) %>% 
    mutate(status="max") # add this column for plottingh later
unique(max_week_filter_perc$class)
```
For visualization, we introduce Na in days that have missing data. 

```{r}
date <- tibble(date = seq(min(data_week_filter$date), max(data_week_filter$date), by="week"))
data_week_filter <- data_week_filter %>%
    select(-c(counts, tot_counts)) %>% 
    # get a table of count species => introduce 0 when species is not present a day
    pivot_wider(names_from=class, values_from = perc, values_fill=0) %>% 
    # add NAs for days without any sampling 
    right_join(date) %>%
    arrange(date) %>% 
    # get back to a long table with the maximum counts and attributed range 
    pivot_longer(Akashiwo:Coscinodiscus, names_to="class",values_to="perc")
```

We add an extra column for the status of each week

```{r}
data_week_filter <- left_join(data_week_filter, max_week_filter_perc)
```
Now we can keep only those maximum classes and an extra one for *Cerataulina bergonii*, update their name to be more representative than what they are called in the database and plot them. 

```{r}
# filter only the max classes and rename them (could be coded better)
toplot = data_week_filter %>% 
  filter(class %in% c(unique(max_week_filter_perc$class),"Cerataulina bergonii")) %>% 
  mutate(class = replace(class,class == "group chaetoceros straight", "Chaetoceros straight")) %>%
  mutate(class = replace(class,class == "group chaetoceros single cell", "Chaetoceros single")) %>%
  mutate(class = replace(class,class == "Protozoa", "Pointed nanophytoplankton")) %>%
  mutate(class = replace(class,class == "Skeletonema", "Skeletonema sp.")) %>%
  mutate(class = replace(class,class == "square unknown", "Square nanophytoplankton")) %>%
  mutate(class = replace(class,class == "Leptocylindrus", "Leptocylindrus sp.")) %>%
  mutate(class = replace(class,class == "unknown2", "Round nanophytoplankton")) %>%
  mutate(class = replace(class,class == "Thalassiosira", "Thalassiosira sp."))

# rename also the max table 
max_week_filter_perc <- max_week_filter_perc %>% 
    mutate(class = replace(class,class == "group chaetoceros straight", "Chaetoceros straight")) %>%
  mutate(class = replace(class,class == "group chaetoceros single cell", "Chaetoceros single")) %>%
  mutate(class = replace(class,class == "Protozoa", "Pointed nanophytoplankton")) %>%
  mutate(class = replace(class,class == "Skeletonema", "Skeletonema sp.")) %>%
  mutate(class = replace(class,class == "square unknown", "Square nanophytoplankton")) %>%
  mutate(class = replace(class,class == "Leptocylindrus", "Leptocylindrus sp.")) %>%
  mutate(class = replace(class,class == "unknown2", "Round nanophytoplankton")) %>%
  mutate(class = replace(class,class == "Thalassiosira", "Thalassiosira sp."))
```

```{r}
# plot
ggplot(toplot, aes(x=date, y=perc, fill=factor(class, levels=c("Pointed nanophytoplankton","Skeletonema sp.", "Chaetoceros straight", "Square nanophytoplankton","Leptocylindrus sp.","Chaetoceros single","Round nanophytoplankton" ,"Thalassiosira sp.","Cerataulina bergonii")))) + 
    geom_area(size=1, alpha=0.5) + 
    geom_col(data=max_week_filter_perc) + 
    facet_wrap(.~factor(class, levels=c("Pointed nanophytoplankton","Skeletonema sp.", "Chaetoceros straight", "Square nanophytoplankton","Leptocylindrus sp.","Chaetoceros single","Round nanophytoplankton" ,"Thalassiosira sp.","Cerataulina bergonii"))) + 
    scale_fill_manual(values = wes_palette("Zissou1", n = 9, type="continuous")) + 
    scale_x_datetime(date_breaks = "4 months",date_labels="%b\n%Y") + 
    theme_ipsum(axis_title_just = "m",axis_title_size = 10) + 
    theme(legend.position="none",
          text = element_text(family="Lato", size=10),
          panel.grid.major.y=element_blank(),
          panel.grid.minor=element_blank(),
          panel.grid.major.x = element_line(linetype = "dashed", color="grey94")) +
    labs(x = "", y="Percentage of the community (%)")

ggsave(filename="F10_phyto.png", path = "../figures/", device='png', dpi=300, width=12,
       height=7,unit="in")
```

