---
title: "Time series of the morphology" # title of the notebook
author: "Virginie Sonnet \nvirginie_sonnet@uri.edu \nhttps://github.com/VirginieSonnet" # name, email and github with links
date: "`r format(Sys.Date(), '%B %dth, %Y')`" # date of today (change th if today is 1st, 2nd or 3rd)
output:
  # specific format from the R package "rmdformats"
  rmdformats::readthedown:
    lightbox: TRUE # allows to click on an image to zoom in the HTML document
    highlight: pygments # color code for code (default is also a good option)
    toc_depth: 5 # level of headers to include in the table of contents 
    # code_folding: "hide" # do not show the code (can access it with a button)
    # use_bookdown: TRUE # automatic section numbering 
    # toc_float: TRUE # for regular markdowns, allow the toc to not be on top but on the side
---

```{=html}
<style>
body {
font-family: "LM Roman"}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

```{r}
library(lubridate)
library(data.table)
library(factoextra)
library(extrafont)
loadfonts()
library(tidyverse)
library(cowplot)
library(castr) # moving average => https://github.com/jiho/castr
library(scales)
library(grid)
library(hrbrthemes)
```

## Introduction

This code looks at the hourly aggregated morphological coordinates: time series, distribution and time series with main contributors.

## Morphological time series

```{r}
# aggregated PC coordinates 
morpho <- read_csv("../data/pca_coord_summary.csv") 
```

Now that we have our aggregated data, we want to regularize the time step to accurately plot the missing values.

```{r}
date <- tibble(date = seq(min(morpho$date), max(morpho$date), "hour"))
morpho <- left_join(date, morpho) 
```

In order to make the plot easier to interpret, we are using a moving average to smooth the time series. We use a window of 60 values (30 before and after) and smooth over the time series 5 times.

```{r}
morpho_mean <- morpho %>% 
  select(date, ends_with("mean") & -starts_with("conc")) %>%
  pivot_longer(PC1_mean:PC4_mean, names_to="pc",values_to="values") %>% 
  mutate(pc = str_remove(pc,"_mean")) %>% 
  group_by(pc) %>% 
  mutate(mavg = slide(values, k=60, mean, n=5, na.rm=TRUE)) %>% 
  mutate(mavg = replace(mavg, is.na(values),NA))
```

Now we can plot the time series.

```{r, fig.height=5.75}
# axis labels
labels <- c("PC1- Size", "PC2 - Width", "PC3 - Regularity", "PC4 - Chains") 


# plot 
p <- morpho_mean %>% ggplot(aes(x=date,y=values)) +
  geom_hline(yintercept = 0, size = 0.5, color = "darkslateblue", linetype="dotted") +
  geom_point(alpha=0.1,size = 0.5,col="grey") +
  geom_path(aes(x=date, y=mavg, col=pc), size=0.9) + 
  # theme
  facet_wrap(.~pc, scale="free_y",strip.position = "left", nrow=4,
             labeller = as_labeller(c("PC1"=labels[1],"PC2" = labels[2], "PC3" = labels[3],"PC4" = labels[4]))) + 
  theme_minimal() + 
  theme(text = element_text(size = 15, family = "Lato"),
        axis.title.y = element_text(size = 12, family = "Lato")) +
  theme(panel.grid.major.x = element_line(size = 0.5, linetype = 'dashed', color="grey77"),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.line.y.left = element_line(),
        axis.ticks.y.left = element_line(),
        strip.placement = "outside",
        legend.position = "none") + 
  # labels 
  labs(x = "", y = "") +
  # axis 
  scale_x_datetime(expand=c(0,0), date_breaks="4 months", date_labels="%b\n%Y") 

```

```{r,fig.height=5.75, warning=FALSE}
p
ggsave(filename="F6_ts_morpho.tiff", path = "../figures/", device="tiff", dpi=300)
```

## Main contributors time series

Let's plot how the different variable contribute to each axis. First, we need to convert all of them from pixels to microns (\~ 3.4 pixels/micron) and to assign one principal component to each.

```{r}
# load the data 
raw <- read_tsv("../data/raw_morpho.csv")

# convert to um
raw <- raw %>% 
  # average per hour 
  mutate(date = round_date(date, "hour")) %>% 
  group_by(date) %>% 
  summarize_all(list(~mean(.,na.rm=TRUE))) %>% 
  # convert to um, um2 and um3 (the extent and area.perim ratios)
  mutate_at(c("min.ax","perim","maj.ax","med.dist","hflip"), ~./3.4) %>% 
  mutate_at(c("area","area.perim2"), ~./(3.4*3.4)) %>% 
  mutate_at(c("vol"),~./(3.4*3.4*3.4)) %>%
  # duplicate the volume column to use it for both PC1 and 2 
  mutate(vol2=vol)

# regularize timespet
date <- tibble(date=seq(min(raw$date),max(raw$date),"hour"))
raw <- left_join(date, raw)

# attribute the PC they contribute to 
raw <- raw %>% 
  pivot_longer(area:vol2, names_to="var", values_to="values") %>% 
  mutate(pc = case_when(var %in% c("vol","area","perim","maj.ax") ~ "PC1",
                        var %in% c("area.perim","min.ax","vol2","area.perim2") ~ "PC2",
                        var %in% c("extent","perim.maj","hflip") ~ "PC3",
                        var %in% c("med.dist","nb.blobs") ~ "PC4")) %>% 
  mutate(var=ifelse(var == "vol2","vol",var)) %>% 
  # add a slide curve 
  group_by(pc,var) %>% 
  mutate(mavg = slide(values, k=60, mean, n=5, na.rm=TRUE)) %>% 
  mutate(mavg = replace(mavg, is.na(values),NA))
```

Now we can add the morphological averages

```{r}
morpho_mean <- add_column(morpho_mean,var=".PCmorpho")
raw.morpho <- bind_rows(raw,morpho_mean)
```

And the last step is to plot.

```{r}
# PC1
p1 <- raw.morpho %>% 
  filter(pc=="PC1") %>%  
  ggplot(aes(x=date,y=values)) +
  geom_hline(yintercept = 0, size = 0.5, color = "darkslateblue", linetype="dotted") +
  geom_point(alpha=0.1,size = 0.5,col="grey") +
  geom_path(aes(x=date, y=mavg, col=ifelse(var==".PCmorpho",1,0)), size=0.9) + 
  # theme
  facet_wrap(.~var, scale="free_y",strip.position = "left",nrow=5) + 
  theme_minimal() + 
  theme(text = element_text(size = 12, family = "Lato"),
        axis.title.y = element_text(size = 12, family = "Lato")) +
  theme(panel.grid.major.x = element_line(size = 0.5, linetype = 'dashed', color="grey77"),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.line.y.left = element_line(),
        axis.ticks.y.left = element_line(),
        strip.placement = "outside",
        legend.position = "none") + 
  # labels 
  labs(x = "", y = "",title="PC1 - Size") +
  # axis 
  scale_x_datetime(expand=c(0,0), date_breaks="6 months", date_labels="%b\n%Y")

ggsave(plot=p1, filename="S4_morpho+raw_1.tiff", path = "../figures/", device='tiff', dpi=300)
```

```{r}
# PC2 
p2 <- raw.morpho %>% 
  filter(pc=="PC2") %>%  
  ggplot(aes(x=date,y=values)) +
  geom_hline(yintercept = 0, size = 0.5, color = "darkslateblue", linetype="dotted") +
  geom_point(alpha=0.1,size = 0.5,col="grey") +
  geom_path(aes(x=date, y=mavg, col=ifelse(var==".PCmorpho",1,0)), size=0.9) + 
  # theme
  facet_wrap(.~var, scale="free_y",strip.position = "left",nrow=5) + 
  theme_minimal() + 
  theme(text = element_text(size = 15, family = "Lato"),
        axis.title.y = element_text(size = 12, family = "Lato")) +
  theme(panel.grid.major.x = element_line(size = 0.5, linetype = 'dashed', color="grey77"),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.line.y.left = element_line(),
        axis.ticks.y.left = element_line(),
        strip.placement = "outside",
        legend.position = "none") + 
  # labels 
  labs(x = "", y = "",title="PC2 - Width") +
  # axis 
  scale_x_datetime(expand=c(0,0), date_breaks="6 months", date_labels="%b\n%Y") 
```

```{r}
# PC3
p3 <- raw.morpho %>% 
  filter(pc=="PC3") %>%  
  ungroup() %>% 
  add_row(var=c("zextra1")) %>%  
  ggplot(aes(x=date,y=values)) +
  geom_hline(yintercept = 0, size = 0.5, color = "darkslateblue", linetype="dotted") +
  geom_point(alpha=0.1,size = 0.5,col="grey") +
  geom_path(aes(x=date, y=mavg, col=ifelse(var==".PCmorpho",1,0)), size=0.9) + 
  # theme
  facet_wrap(.~var, scale="free_y",strip.position = "left",nrow=5) + 
  theme_minimal() + 
  theme(text = element_text(size = 15, family = "Lato"),
        axis.title.y = element_text(size = 12, family = "Lato")) +
  theme(panel.grid.major.x = element_line(size = 0.5, linetype = 'dashed', color="grey77"),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.line.y.left = element_line(),
        axis.ticks.y.left = element_line(),
        strip.placement = "outside",
        legend.position = "none") + 
  # labels 
  labs(x = "", y = "",title="PC3 - Regularity") +
  # axis 
  scale_x_datetime(expand=c(0,0), date_breaks="6 months", date_labels="%b\n%Y") 

ggsave(plot=p3, filename="S4_morpho+raw_3.tiff", path = "../figures/", device='tiff', dpi=300)
```

```{r}
# PC4
p4 <- raw.morpho %>% 
  filter(pc=="PC4") %>%  
  ungroup() %>% 
  add_row(var=c("zextra1","zextra2")) %>%  
  ggplot(aes(x=date,y=values)) +
  geom_hline(yintercept = 0, size = 0.5, color = "darkslateblue", linetype="dotted") +
  geom_point(alpha=0.1,size = 0.5,col="grey") +
  geom_path(aes(x=date, y=mavg, col=ifelse(var==".PCmorpho",1,0)), size=0.9) + 
  # theme
  facet_wrap(.~var, scale="free_y",strip.position = "left",nrow=5) + 
  theme_minimal() + 
  theme(text = element_text(size = 15, family = "Lato"),
        axis.title.y = element_text(size = 12, family = "Lato")) +
  theme(panel.grid.major.x = element_line(size = 0.5, linetype = 'dashed', color="grey77"),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.line.y.left = element_line(),
        axis.ticks.y.left = element_line(),
        strip.placement = "outside",
        legend.position = "none") + 
  # labels 
  labs(x = "", y = "",title="PC4 - Chains") +
  # axis 
  scale_x_datetime(expand=c(0,0), date_breaks="6 months", date_labels="%b\n%Y") 

ggsave(plot=p4, filename="S4_morpho+raw_4.tiff", path = "../figures/", device='tiff', dpi=300)

```

## Morphological distribution per season

For this part, we need the full dataset in order to derive the distribution. We will consider that the year 2018 goes from November 2017 to

```{r}
# colnames(full) <- c("id","roi_number","class_id","date","conc","PC1","PC2","PC3","PC4")

# colorpalette
colorpalette <- c(rgb(0,175,187,50,maxColorValue=255),
                  rgb(0,175,187,maxColorValue=255),
                  rgb(0,100,0,50,maxColorValue=255),
                  rgb(0,100,0,maxColorValue=255),
                  rgb(252,78,8,50,maxColorValue=255),
                  rgb(252,78,8,maxColorValue=255),
                  rgb(242,215,73,50,maxColorValue=255),
                  rgb(242,215,73,maxColorValue=255))

full <- fread("../data/full_data_pca.csv") %>% 
  select(-c(id,roi_number,class_id,conc)) %>% 
  rename("PC1" = "1", "PC2"="2","PC3"="3","PC4"="4") %>% 
  # add year 
  mutate(year = case_when(date < "2018-11-01" ~ "2018",
                          date >= "2018-11-01" ~ "2019")) %>% 
  # add season 
  mutate(quarter = quarter(date)) %>% 
  mutate(quarter = case_when(quarter==1 ~ "Winter",
                             quarter==2 ~ "Spring",
                             quarter==3 ~ "Summer",
                             quarter==4 ~ "Fall")) %>% 
  mutate(quarter=factor(quarter, levels=c("Winter","Spring","Summer","Fall"))) %>%
  # attribute a color 
  mutate(col = case_when(quarter == "Winter" & year==2018 ~ "#00AFBB32",
                         quarter == "Winter" & year==2019 ~ "#00AFBB",
                         quarter == "Fall" & year==2018 ~ "#00640032",
                         quarter == "Fall" & year==2019 ~ "#006400",
                         quarter == "Spring" & year==2018 ~ "#FC4E0832",
                         quarter == "Spring" & year==2019 ~ "#FC4E08",
                         quarter == "Summer" & year==2018 ~ "#F2D74932",
                         quarter == "Summer" & year==2019 ~ "#F2D749")) %>%
  mutate(col = factor(col, levels = colorpalette)) %>%
  # format long 
  pivot_longer(PC1:PC4, values_to="values", names_to="pc") 
```

```{r}
full %>%
  ggplot(aes(x=quarter, y=value, fill=col, group=col, color=col)) +
  geom_violin(position=position_dodge(width = 0.9)) + 
  geom_boxplot(width=0.1,position = position_dodge(width =0.9),fill="white") + 
  facet_wrap(.~pc, nrow=2,ncol=2, scales="free",strip.position="left",
             labeller=as_labeller(c("PC1"="PC1 - Size","PC2"=" PC2 - Width","PC3"="PC3 - Regularity","PC4"="PC4 - Colonies"))) +
  stat_compare_means(aes(label = ..p.format..), method="wilcox", size=4, family="Lato") + 
  scale_fill_manual(values=colorpalette)+
  scale_color_manual(values=colorpalette)+
  # theme 
  theme_classic() + 
  theme(text = element_text(size=15, family="Lato"),
        panel.grid=element_blank(),
        legend.position="none",
        strip.background=element_blank(),
        strip.placement="outside") +
  # axis 
  labs(x="",y="",fill="")
```

## Zooplankton

We have data available on **zooplankton** only between November 2018 and March 2019. We can look at the time series during that period with the morphological composites variables and the nutrients, temperature and salinity.

```{r}
# wide format morpho
data <- morpho_mean %>% 
  select(-mavg) %>% 
  pivot_wider(names_from="pc", values_from="values")

# environment 
envt <- read_csv("../data/NBay_hourly.csv", guess_max=1000) %>% 
  select(-c(no3:nh4))
data <- left_join(data, envt)

# zooplankton 
zoo <- read_csv("../data/zooplankton_nicole_flecchia.csv") %>% 
  # rename for easiness 
  select(date, vol_m3=`volume sampled (m3)`, copepod_nauplii="copepod nauplii",
         acartia_copepopids="acartia copepopids",
         a.hudsonica_adults="a. hudsonica adults",a.tonsa_adults="a. tonsa adults",
         other_copepods="other copepods",barnacle_nauplii="barnacle nauplii",
         barnacle_cyprids="barnacle cyprids",hydromedusa="hydromedusa",
         polychete_worms="polychete worms",fish_eggs="fish eggs",
         bivalve:gastropods) %>% 
  # transform in cells/m3
  mutate_at(vars(copepod_nauplii:gastropods), ~./vol_m3) %>% 
  # add 8am to the time so that we can join 
  mutate(date = mdy_hms(paste(date,"08:00:00"))) %>% 
  # sum the total zooplankton and copepods
  mutate(tot_zoo = rowSums(select(.,copepod_nauplii:gastropods)),
         tot_cop = rowSums(select(.,copepod_nauplii:other_copepods))) %>% 
  select(-c(vol_m3:gastropods))
data <- left_join(data,zoo)
```

The zooplankton and environment data are on a weekly basis so we will average per week.

```{r}
data <- data %>% 
  mutate(date = floor_date(date,"week")) %>% 
  group_by(date) %>% 
  summarize_all(list(~mean(.,na.rm=TRUE))) %>% 
  drop_na(tot_zoo) %>% 
  pivot_longer(PC1:tot_cop,names_to="var",values_to="values") %>% 
  mutate(type = case_when(var %in% c("di_no","di_po","si") ~ "Nutrients (uM)",
                          var %in% c("PC1","PC2","PC3","PC4") ~"Morpho",
                          var %in% c("tot_zoo","tot_cop") ~ "Zooplankton \n(cells/m3)",
                          var == "temp" ~ "Temperature (C)",
                          var == "sal" ~ "Salinity",
                          var == "light" ~ "Light (W/m2)")) %>% 
  mutate(var = case_when(var == "di_no" ~ "nitrogen",
                         var == "di_po" ~ "phosphate",
                         var == "si" ~ "silicate",
                         var == "tot_cop" ~ "copepods",
                         var == "tot_zoo" ~ "zooplankton",
                         var == "PC1" ~ "size",
                         var == "PC2" ~ "width",
                         var == "PC3" ~ "regularity",
                         var == "PC4" ~ "chains",
                         TRUE ~ var))
```

Now we can plot them together.

```{r}
data %>% 
  mutate(type=factor(type, levels=c("Morpho","Zooplankton \n(cells/m3)", "Nutrients (uM)","Temperature (C)","Salinity","Light (W/m2)"))) %>% 
  ggplot(aes(x=date, y=values, color=var)) +
  geom_line(size=1) + 
  geom_point(size=1.5) + 
  facet_wrap(.~type, scales="free_y",nrow=6,strip.position="left") + 
  scale_y_comma(breaks = breaks_extended(4)) + 
  scale_x_datetime(date_labels="%b \n%Y") + 
  scale_color_manual(values=c("orange2","slateblue2","lightgoldenrod2",
                              "paleturquoise2","deepskyblue3","firebrick4",
                              "springgreen2","dodgerblue4","gold1","lightcoral","coral3", "darkorchid3")) + 
  theme_bw() + 
  theme(strip.placement="ourside",
        strip.background=element_blank(),
        strip.text = element_text(size=14,family="Lato"),
        axis.text=element_text(size=12,family="Lato"),
        panel.border=element_blank(),
        legend.position="null",
        panel.grid.minor.y=element_blank(),
        legend.text = element_text(size=14,family="Lato"),
        panel.grid.major.x = element_line(color="grey85",size=0.75)) + 
  labs(x="",y="")

ggsave(filename="S5_zoo.svg", path = "../figures/", device='svg', dpi=300,
       width=6.5,height=10,unit="in")
```
